---
title: "R Notebook"
output: html_notebook
---

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. 

```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

Read data
```{r}
library(dplyr)
df1 = read.csv('data/data1.csv')
df2 = read.csv('data/data2.csv')
df3_youth = read.csv('data/data3_youth.csv')
df3_adult = read.csv('data/data3_adult.csv')
df1_filter = df1 %>% select('SUB.AREA','SEASON','AGE.GROUP','WEEK','NUM.INFLUENZA.DEATHS','NUM.PNEUMONIA.DEATHS','NUM.COVID.19.DEATHS','TOTAL.DEATHS')
df2_filter = df2 %>% select('MMWR.Year','MMWR.Week','All.Cause','Natural.Cause','Septicemia..A40.A41.','Malignant.neoplasms..C00.C97.','Diabetes.mellitus..E10.E14.','Alzheimer.disease..G30.','Influenza.and.pneumonia..J09.J18.','Chronic.lower.respiratory.diseases..J40.J47.','Other.diseases.of.respiratory.system..J00.J06.J30.J39.J67.J70.J98.','Nephritis..nephrotic.syndrome.and.nephrosis..N00.N07.N17.N19.N25.N27.','Symptoms..signs.and.abnormal.clinical.and.laboratory.findings..not.elsewhere.classified..R00.R99.','Diseases.of.heart..I00.I09.I11.I13.I20.I51.','Cerebrovascular.diseases..I60.I69.','COVID.19..U071..Multiple.Cause.of.Death.','COVID.19..U071..Underlying.Cause.of.Death.')
df3_youth_filter = df3_youth %>% select('Geography_Name','Demographic_Name','Month_Week','n_unweighted','n_weighted','Influenza_Season','Legend_Label') %>% filter(Geography_Name !='National')
df3_adult_filter = df3_adult %>% select('Geographic_Name','Demographic_Name','Month_Week','n_unweighted','Influenza_Season','Legend') %>% filter(Geographic_Name !='National')
head(df1_filter)
df2_filter
head(df3_youth_filter)
head(df3_adult_filter)
```
For dataset 1, we have the following that have missing values:
```{r}
df1_filter %>% filter(NUM.INFLUENZA.DEATHS == 'Insufficient Data')
```
For dataset 2, we have missing values as shown below:
```{r}
head(df2_filter[!complete.cases(df2_filter),] %>% select(!c('MMWR.Year','MMWR.Week')))
```
For dataset 3, we have missing values as follows:
```{r}
head(df3_youth_filter[!complete.cases(df3_youth_filter),])
head(df3_adult_filter[!complete.cases(df3_adult_filter),])
```


